Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/elife.tex
Missed fig8 label
Missed fig8 supplement references

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated $\R[dev]{A}{j}$ with $x_j$ at the start of later trials across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial. Similar behavior was also seen in the \msc{} sequence, as shown in \FIGSUPP[8]{correlation-decay}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-synergy}
    \caption{%
        In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:8}
    % 
    \figsupp[The equivalent figure in the ablated model with only STD.]{%
        The equivalent of \FIG{synergy} in the ablated model with only STD.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-synergy-noTA}}\label{figsupp:noTA}
    %
    \figsupp[The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.]{%
        The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.
        Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.
    }
    {\includegraphics[width=8cm]{sfig-STD-correlation-decay}}\label{figsupp:correlation-decay}
\end{figure}

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{threshold-initial} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}B, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}C, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \FIGSUPP[8]{noTA}).

>>>>

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated $\R[dev]{A}{j}$ with $x_j$ at the start of later trials across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial. Similar behavior was also seen in the \msc{} sequence, as shown in \FIGSUPP[synergy]{correlation-decay}.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-synergy}
    \caption{%
        In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:synergy}
    % 
    \figsupp[The equivalent figure in the ablated model with only STD.]{%
        The equivalent of \FIG{synergy} in the ablated model with only STD.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-synergy-noTA}}\label{figsupp:noTA}
    %
    \figsupp[The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.]{%
        The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.
        Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.
    }
    {\includegraphics[width=8cm]{sfig-STD-correlation-decay}}\label{figsupp:correlation-decay}
\end{figure}

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{threshold-initial} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}B, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}C, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \FIGSUPP[synergy]{noTA}).
